Towards a Wide-Coverage Grammar for Swedish\\ Using GF

### 2012-01-18 doc/report/report.tex
varios small fixes

\abstract{
This thesis describes work towards a wide-coverage grammar for
parsing and generating Swedish text.
We do this by using the dependently typed grammar formalism GF,
a functional programming language specialized at describing grammars.
The idea is to 
%The goal of this thesis is to enable open-domain parsing for Swedish
%by using the GF \\grammar formalism,
%by %The idea has been to% 
combine existing language resources with
new techniques, with an aim to achieve a parser
for unrestricted Swedish.
To reach this goal, problems of computational as well as linguistic nature had
to be solved.  
The work includes the development of the grammar -- \;
%in the GF formalism
to identify and formalize 
grammatical constructions %specific to and
frequent in Swedish -- as well as methods for importing a large-scale lexicon and for evaluating
the parser. % have been investigated and formalised in the functional
%programming language.
%~~achieve a covering Swedish~~ grammar implemented
%and the aim to reaching our goal;
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
We present the methods and technologies used 
and discuss the advantages and problems of using GF
for modeling large-scale grammars.
We further discuss how our long-term goal can be reached
%the possibilities
%of reaching our long-term goal % -- to accomplish a parser for unrestricted Swedish -- 
by combining our rule-based grammar with statistical means.
%We compare and discuss .
%and for obtaining a system for analysing written text.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
into the GF notation and an extended Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in the GF resource library, and now
also covers constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

% check references! lncs is good maybe. lnai
\newpage
%\pagestyle{empty}
%\cleardoublepage
\tableofcontents
%
\addtocontents{toc}{\protect\thispagestyle{empty}}
%\pagestyle{empty}
%\cleardoublepage
\newpage
%\pagenumberingstyle{plain}
\setcounter{page}{1} %reset the page counter
\chapter{Introduction}
\thispagestyle{plain}
% more intro to gf. dependently typed...high-level formalism
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism.
%with facilities for multilinguality. 
It is based on Martin-Löf type theory which allows
reasoning within the programming language.
GF has strong support for multilinguality and has so far been used successfully for 
controlled languages \cite{cnl}. Recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
% >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
% --> sentence about connection GF - grammar based parser.
%This project focuses on implementing a wide-coverage Swedish grammar, which
%describes a parser accepting all sentences that can be expressed by the grammar.
Parsing, that is the task of automatically identifying morphological and
syntactical structures, is receiving increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being further developed. Our
goal is
to create an open-source Swedish grammar from which we derive a parser
accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The shell can be tried out online 
together with an interactive GF editor %and the GF shell 
\footnote{http://www.grammaticalframework.org/demos/gfse/}.
\thispagestyle{plain}
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}
Figure \ref{fig:shell} shows commands for parsing, where the results are shown as
in figure \ref{fig:gftree1}. 
GF uses Graphviz\footnote{http://www.graphviz.org/}
to visualize the trees and the given commands (fig. \ref{fig:shell}) specify that the
output format should be PDF-files and that these should be opened by the program
Evince. The user can choose to see the \textit{parse tree} (fig. \ref{pic:gfStree})
or the \textit{abstract tree} (fig. \ref{pic:gfAtree}). Abstract trees are more verbose and show all
functions and types used for parsing the sentence. 
\begin{figure}[h]
\centering
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
\vspace{10mm}

Parse trees on the other hand show only the types assigned to the words and phrases.
Information about tense, polarity etc. which are explicitly given in the abstract tree are
not reproduced in the parse tree. Hence, parse trees do not give complete representations
but model the parse results in a transparent manner. For our example, the definiteness and
number of the noun `\emph{katten}' is shown as \verb-DetQuant DefArt NumSg- in
the abstract tree while the parse tree only show that a the noun phrase consists of one noun. 
% is used but not in the parse tree.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{`the'},
and the determiner, the first argument to the function \verb-DetCN-, is
therefore shown in the parse tree.\\

>>>>

\abstract{
This thesis describes work towards a wide-coverage grammar for
parsing and generating Swedish text.
We do this by using the dependently typed grammar formalism GF,
a functional programming language specialized at describing grammars.
The idea is to 
%The goal of this thesis is to enable open-domain parsing for Swedish
%by using the GF \\grammar formalism,
%by %The idea has been to% 
combine existing language resources with
new techniques, with an aim to achieve a parser
for unrestricted Swedish.
To reach this goal, problems of computational as well as linguistic nature had
to be solved.  
The work includes the development of the grammar -- \;
%in the GF formalism
to identify and formalize 
grammatical constructions %specific to and
frequent in Swedish -- as well as methods for importing a large-scale lexicon and for evaluating
the parser. % have been investigated and formalised in the functional
%programming language.
%~~achieve a covering Swedish~~ grammar implemented
%and the aim to reaching our goal;
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
We present the methods and technologies used 
and discuss the advantages and problems of using GF
for modeling large-scale grammars.
We further discuss how our long-term goal can be reached
%the possibilities
%of reaching our long-term goal % -- to accomplish a parser for unrestricted Swedish -- 
by combining our rule-based grammar with statistical methods.
%We compare and discuss .
%and for obtaining a system for analysing written text.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
into the GF notation and an extended Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in the GF resource library, and now
also covers constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

% check references! lncs is good maybe. lnai
\newpage
%\pagestyle{empty}
%\cleardoublepage
\tableofcontents
%
\addtocontents{toc}{\protect\thispagestyle{empty}}
%\pagestyle{empty}
%\cleardoublepage
\newpage
%\pagenumberingstyle{plain}
\setcounter{page}{1} %reset the page counter
\chapter{Introduction}
\thispagestyle{plain}
% more intro to gf. dependently typed...high-level formalism
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism.
%with facilities for multilinguality. 
It is based on Martin-Löf type theory which allows
reasoning within the programming language.
GF has strong support for multilinguality and has so far been used successfully for 
controlled languages \cite{cnl}. Recent experiments have  however showed
that it is also possible to use the framework for parsing free language \cite{patent}.
% >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
% --> sentence about connection GF - grammar based parser.
%This project focuses on implementing a wide-coverage Swedish grammar, which
%describes a parser accepting all sentences that can be expressed by the grammar.
Parsing, that is the task of automatically identifying morphological and
syntactical structures, is receiving increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being further developed. Our
goal is
to create an open-source Swedish grammar from which we derive a parser
accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The shell can be tried out online 
together with an interactive GF editor %and the GF shell 
\footnote{http://www.grammaticalframework.org/demos/gfse/}.
\thispagestyle{plain}
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}
Figure \ref{fig:shell} shows commands for parsing, where the results are shown as
in figure \ref{fig:gftree1}. 
GF uses Graphviz\footnote{http://www.graphviz.org/}
to visualize the trees and the given commands (fig. \ref{fig:shell}) specify that the
output format should be PDF and that these files should be opened by the program
Evince. The user can choose to see the \textit{parse tree} (fig. \ref{pic:gfStree})
or the \textit{abstract tree} (fig. \ref{pic:gfAtree}). Abstract trees are more verbose and show all
functions and types used for parsing the sentence. 
\begin{figure}[h]
\centering
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\caption{Abstract tree and parse tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
\vspace{10mm}

Parse trees on the other hand show only the types assigned to the words and phrases.
Information about tense, polarity etc., which are explicitly given in the abstract tree, are
not reproduced in the parse tree. Hence, parse trees do not give complete representations
but model the parse results in a transparent manner. For our example, the definiteness and
number of the noun `\emph{katten}' is shown as \verb-DetQuant DefArt NumSg- in
the abstract tree while the parse tree only shows that the noun phrase consists of one noun. 
% is used but not in the parse tree.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{`the'},
and the determiner, the first argument to the function \verb-DetCN-, is
therefore shown in the parse tree.\\
